Hire Ed: Job Market Dynamics for Tenure-Track Faculty Positions in Archaeology

### 2025-01-08 analysis/paper/article.qmd
First round of Jess edits to body of the paper.

title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 900
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Academic careers are frequently sought by archaeology graduate students. Job listing websites often serve as the first place for these students when seeking academic positions. We examined tenure-track job advertisements over the past decade to gain insights into the academic job market for archaeologists. Using data from the community-edited Academic Jobs Wiki for Archaeology, we examine changes in the academic job market over time. We studied the text of 431 job ads posted from 2013-2023. Our analysis focuses on shifts in archaeological topics, methods and geographic regions mentioned in job ads. We investigate whether the burden on applicants has changed over time: do institutions request more information and documents from applicants at the initial stages of application, compared to a decade ago? We assess the influence of socio-political factors on the changing focus of research topics in the field. This research aims to assist current and future archaeology students and graduates in better understanding the job market and the requirements of employers, thereby aiding them in effectively preparing for their applications for positions in archaeology.
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

```{r}
#| label: fig-show-basic-plots
#| fig-cap: "A: total number of job ads posted to the Academic Jobs Wiki for Archaeology in each year, with coloured sections showing the proportion of jobs by title and rank. B: Ratio of tenure-track to non-tenure-track positions over time."

# show figure from PNG file
knitr::include_graphics(here("analysis",
            "figures", 
            "fig-panel-per-year.png"))
```

@@

@@

The academic job market is a source of uncertainty, especially for job seekers, as they look for ads that are well-matched to their niche of skills and experience that have taken years to cultivate and refine. Hiring committees also face uncertainty as they strive to equitably sift through sometimes hundreds of applicants in the hope of finding a candidate that meets all of their needs and helps to realise their visions of their program and the discipline. For many academics there is a perception that the last two to three decades have seen the job market become increasingly difficult for job seekers. In many fields it is common to hear that there are fewer jobs available relative to the number of PhD graduates, higher numbers of short-term temporary positions relative to secure tenure-track positions, and increasingly complex application requirements with specific statements (e.g. teaching, research, and diversity) that need to be tailed for each application, making the process of applying for jobs a full-time job of its own.

In anthropology in the US the number of doctoral anthropology graduates has increased by about 70% over the past 30 years, but the number of new faculty positions has not increased proportionally [@speakman2018choosing]. Among biological anthropologists, @Passalacqua_2018 found a ratio of 0.81 PhDs to job academic advertisements per year, concluding that academic positions in biological anthropology are barely at sustainable levels. This echos findings from other fields, for example in biomedical sciences where there is one tenure-track position in the US for approximately every 6.3 PhD graduates [@Ghaffarzadegan_2015]. PhD graduates who do not go directly into a tenure-track faculty position after graduation often go into a series of short-term appointments as part-time or limited contract instructors, also known as adjunct or contingent faculty. Rising numbers of contingent faculty has long been a concern in US higher education [@Trevithick2010]. Data from the American Association of University Professors' 2021–22 faculty survey indicate that more than 60% of faculty positions in US universities were held by non-tenure track full time or part time contingent faculty members [@colby2022annual]. Contingent faculty positions are precarious because they provide low or no health and retirement benefits, little or nothing in the way of professional development, and lower salaries relative to tenure-track positions. PhD graduates who have a sequence of short-term academic positions are often disadvantaged financially, due to low compensation combined with expensive frequent relocation, socially due to isolation from family and community, and professionally due to being ineligible for many decision-making roles at the universities they work at [@platzerAcademicPrecarityAmerican2018].

Studies of who holds tenure-track faculty positions in archaeology at US universities, shows that success in job applications is strongly determined by where applicants get their PhD from. Data from the 2014–2015 AnthroGuide publication of the American Anthropological Association shows that just ten out of over 100 US graduate programs produced over 30% of the graduates hired into tenure-track faculty positions [@speakman2018choosing]. Similarly, a network analysis of 1,918 faculty holding tenured or tenure-track positions at PhD-granting anthropology programs in the US in 2015 (including 506 archaeologists) found that the just fifteen graduate programs produced 52.8 percent of tenured and tenure-track positions [@kawaSocialNetworkUS2019]. This network analysis further found that programs with large endowments and with faculty who hold prestigious awards and are widely cited by other scholars produce the majority of tenured and tenure-track faculty in anthropology. This pattern of inequality is not unique to archaeology [@wapmanQuantifyingHierarchyDynamics2022a], but these studies draw attention to the paradox of a discipline that has many members committed to fighting social inequalities, and yet continues to reproduce its own systematic inequalities and hierarchies through hiring practices that favour an elite minority of graduates with prestigious institutional affiliations.

One way that some hiring committees are tackling these biases towards prestigious institutions is by providing detailed instructions to applicants on how to prepare their application materials so that the hiring committee can focus on evaluating and ranking the specific accomplishments of candidates across common categories, rather than ranking based on prestige signals in a CV, such as a name of the applicant's graduate program. This has resulted in job ads that are often highly prescriptive in the types of documents that applicants should submit. For example, in addition to a cover letter and CV, job ads in many fields now require applicants to submit short statements about their teaching and research record and plans. In a comparison of job ads from 1999–2000 and 2019–2020 published in *Anthropology News* @HelloTristesTropes2021 noticed a trend toward increasingly asking for more materials from applicants. They found that twice as many 2019–2020 job ads requested writing samples compared to 1999–2000, and nearly four times as many requested statements of teaching philosophy. In their review of the 'worst job ads of 2021' @dennisWorstAnthroJob2022 reported two job ads with particularly extreme requirements, requesting nine and thirteen documents from applicants. This high burden on applicants disproportionately favors people with more time and financial resources to prepare the required materials.

A notable requirement that has emerged in recent years is a diversity statement, where the applicant describes their knowledge of, prior contributions to, and future goals for advancing diversity, equity, inclusion. This requirement has been much-debated largely due to an experiment in 2016-2022 at several University of California campuses that used diversity statements as the first cut for selecting candidates for tenure-track faculty positions [@soucek2021diversity]. Only candidates that scored highly on their diversity statements would have the rest of their application evaluated. This experiment generated intense and widespread public debate about merits and risks of requiring and using diversity statements. These debates drew attention to the challenges of hiring faculty from underrepresented minorities, and resulted a variety of approaches to evaluating diversity, including some universities dropping requirements for diversity statements in job applications [@guidenAreDiversityStatements].

@@

Our primary data source is the Archaeology Academic Jobs Wiki. Originating in 2007, this is a set of freely accessible web pages that anyone can edit (anonymously or with a free user account) hosted by Fandom, a for-profit company. The Archaeology pages are part of the Academic Jobs Wiki, which coordinates similar collaboratively-edited resources for around 40 academic disciplines. The coordinators and contributors are nearly all anonymous or pseudonymous. Typically contributors copy and paste the text of job ads from other sources, such as the *Chronicle of Higher Education*, *Higher Ed Jobs*, and university websites, into the wiki, collecting ads originally posted in numerous different locations. Other contributors then edit the web page to add comments below an ad to share relevant information based on their experience in applying for that position, such as a tally of how many people have applied, the dates of events such as requests for more materials, interviews, offer made, rejection notices, etc. Contributors also edit the page to ask and answer questions about the positions and the application process. These comments make the Academic Jobs Wiki a unique resource for timely and specific information for job-seekers about positions they are interested in, and one of the most important internet resources for the academic job market. Because of its reputation for aggregating ads from diverse sources and rapidly-updated information that is not available elsewhere, the Academic Jobs Wiki has a large community of users that keep it updated and accurate has become an authoritative data source for studies of hiring trends in academia [e.g. @Passalacqua_2018; @musial2018five] and a widely recommended resource [e.g. @lightfoot2021preparing].

For each tenure-track job advertised on the Archaeology Academic Jobs Wiki during 2013-2023 we read the text and recorded into a Google form the name of the hiring institution, the title of the position, and exact words and phrases from the ad about the topical, geographic, and methods foci on the position. The topical focus is what we understood as the primary intellectual focus of the position. The geographic focus is the region of the world that the ideal candidate has scholarly expertise on. The methods focus is the data-generating sub-field of archaeology that is mentioned in the ad. We recorded the type and number of documents requested in each ad (e.g. cover letter, CV, statements on research, teaching, diversity, syllabi, course descriptions, writing samples, transcripts) and how many names/letters of recommenders were requested in the ad.

After completing primary data collection, we studied the topical, geographic, and methods foci of each ad and collaboratively and manually reduced the variation in the raw data into 10-15 categories appearing in 20 (for topics and geography) or 10 (for methods) or more job ads to simplify analysis and visualisation. This means that some topics, such as gender (mentioned in `r n_jobs_mentioning_gender` ads) do not appear in our results because of their rarity in the job ads. Full details of the category reduction, showing the mapping between phrases in the job ads and our categories, are in our Supplementary Materials. Our topic categories were: American archaeology, Ancient Europe and Mediterranean, Archaeological science, Archaeological theory, Biological anthropology, Complex societies, Digital archaeology, Environmental archaeology, Evolutionary anthropology, Indigenous and Historical archaeology, North Mesoamerican Archaeology, Pleistocene archaeology, and Public archaeology Our geographic categories were: Africa, Americas, Asia & India, Canada & Arctic, Europe, Mediterranean, Meso- & South America, Near East, Oceania, Midwest US, Northeastern US, Southeast US, Southwest US, and Western US. Our methods categories were: Archaeobotany, Archaeometry, Bioarchaeology, Ceramic analysis, Computational and Digital archaeology, Geoarchaeology, Landscape analysis, Lithic analysis, Material culture analysis, and Zooarchaeology. Ads could have multiple or none of these three foci, and some of the foci overlap. Some topics include geographic regions because this is how they are typically understood by archaeologists. For example Mesoamerican archaeology is understood to refer to a specific time period and geographic region. Similarly, digital archaeology we recorded as both a method (when a job ad has a clearly distinct topical focus, such as historic archaeology) and a topic (when there is no other topics mentioned in the job ad). While these overlaps can make the data challenging to interpret, in our view it reflects the complex realities of how search committees express their needs in searching for new faculty, and is insightful in how it reveals intersections between different foci.

@@

We collected data from `r total_number_of_ads_in_our_sample` ads for tenure-track jobs in archaeology posted during 2013-2023, and focus our analysis here on the `r total_number_of_us_ads_in_our_sample` ads for positions at US universities. @fig-show-basic-plots shows the count of ads for each year. Assistant Professor jobs are consistently the most common title and rank, and open rank or full professor are the least frequent. The ratio of tenure-track to non-tenure track positions is generally well above one. Only 2013-2014 had more non-tenure track positions than tenure track, which was followed by an upward trend peaking at 2018-2019 and the declining again into the present.

@@

```{r}
#| label: fig-show-map-of-hiring-institution-and-cc
#| fig-cap: "A: Frequency of hiring institution by Carnegie classification. B: Inset shows map of the United States showing the count of tenure-track job ads posted by all insititutions in each state during 2013-2023"

@@

Panel B of @fig-show-map-of-hiring-institution-and-cc shows the geographic distribution of the hiring institutions. California posted almost twice as many job ads as the next most active states. After California, the states that posted the most ads during 2013-2023 include New York, Texas, and Pennsylvania, and Florida. These top five states also correspond to the top five most populous US states, indicating that rates of hiring is approximately proportional to population density. These top five states for job ads are also the top five states with highest number of degrees awarded in Anthropology (U.S. Department of Education nd). Similarly, the lowest counts of job ads were observed in states with the lowest populations: North Dakota, South Dakota, Alaska, and Nebraska. No institutions in Montana posted a job ad during this period. The implication here is that job-seekers who are able to relocate to populous areas will have more employment options.

@@

@@

We recorded all geographic regions mentioned in the text where the successful applicant should have expertise and be research active. Our analysis focuses on those locations mentioned in 20 or more ads. Overall, American locations dominate. Panel A of @fig-show-geo-trends shows that a single region of the US, the Southwest, occurs in more job ads than every other part of the world except for the Mediterranean. The Southwest includes Arizona and New Mexico, with portions of California, Colorado, Nevada, Oklahoma, Texas, and Utah. It is archaeologically significant as the home of the Ancestral Pueblo, Hohokam, and Mogollon peoples who practiced irrigation agriculture and lived in relatively large settlements, compared to other regions of the US. The area was later occupied by the Navajo, Ute, Southern Paiute, Hopi and Zuni, also in relatively large numbers. The Mediterranean is prominent because it is the location that is often mentioned in job ads focused on classical archaeology (i.e. archaeology of Bronze and Iron Age Italy and Greece).

Demand for jobs focusing on the Americas is generally high over time, with a peak in 2019-2020, and a decrease since then. Demand for jobs focusing on Africa was very low until 2019-2020, peaking in 2020-2021. The proportion of ads with a geographic focus on the Mediterranean has varied substantially, peaking at 2016 and at its lowest in 2019, showing an inverse pattern of the Americas. Asia and India, the Near East and Europe are consistently rare as a geographical focus in job ads. Asia and India, Africa and the Americas appear correlated with each other, while the Near East and Mediterranean are correlated in an opposite trend.

@@

Landscape archaeology, encompassing GIS and remote sensing, has remained prominent compared to other methods [@fig-show-metho-trends]. Methods focused on a specific element of the archaeological record, such as Lithic analysis, Zooarchaeology and Ceramics are among the least frequently mentioned in job ads. Instead we see the more popular methods here are ones that are relevant to multiple elements of the archaeological record (e.g. Archaeobotany encompasses macroscopic and microscopic plant remains; Bioarchaeology may include skeletal analysis, isotopes, proteins, etc.).

Landscape archaeology, although dominant, has fluctuated over the years and has been on a downtrend since 2018-2019. Computational and digital archaeology is the second most represented method, showing an overall increasing trend, particularly since 2020-2021. Archaeobotany shows a strong cyclical trend, with a rise and fall and rise again over our study period. Archaeometry and Geoarchaeology have maintained a relatively low but steady presence in job ads, peaking in 2017-2018 and 2018-2019 and declining thereafter. Lithic analysis and Zooarchaeology are also mentioned relatively infrequently in job ads and show an inverse correlation with each other after 2018-2019.

@@

@@

The most frequently mentioned topic in the job ads in Environmental archaeology [@fig-show-topi-trends], which is our category for phrases found in the text of ads such as human-environmental dynamics, interaction between humans and their environments, environmental change, climate change, historical ecology, ecological knowledge,human ecology, and ecological systems. Public archaeology is the second most frequent topic overall, this includes phrases such as cultural resource management, cultural heritage, heritage studies, museum studies, human rights, community engaged, historic preservation; social justice, community-based, repatriation, and community-engaged archaeology. The least frequent topics are Pleistocene archaeology, e.g. human origins; hunter-gatherer archaeology, and digital archaeology.

In the years 2019-2020 and 2020-2021 there are striking changes in the popularity of topics in job ads. Indigenous and historical archaeology (which includes archaeology of enslaved people) became the most popular topic at this time, rising from being one of the least popular topics from 2012-2017. Conversely, archaeological science, which was popular during 2012-2017, was rarely mentioned in job ads during 2019-2021. Ancient Europe and the Mediterranean, which was consistently one of the least frequently mentioned topics, virtually disappeared from job ads during 2019-2021. Mentions of complex societies in ads consistently decrease over the study period, while mentions of public archaeology consistently increase. The topic of Environmental Archaeology remains consistently high over time.

>>>>

title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 900
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Many archaeology graduate students pursue advanced degrees in the hope of landing an academic career. Job listing websites often serve as the first place for these students when seeking academic positions. We examined tenure-track job advertisements over the past decade to gain insights into the academic job market for archaeologists. Using data from the community-edited Academic Jobs Wiki for Archaeology, we examine changes in the academic job market over time. We analyzed the text of 431 job ads posted from 2013--2023. Our analysis focuses on shifts in archaeological topics, methods and geographic regions specified in archaeological job ads. We also investigate whether the labor burden for applicants has changed over time: do institutions request more information and documents from applicants at the initial stages of application, compared to a decade ago? We assess the influence of socio-political factors on the changing focus of research topics in the field. This research aims to assist current and future archaeology students and graduates in better understanding the job market and the requirements of employers, thereby aiding them in effectively preparing for their applications for positions in archaeology.
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

```{r}
#| label: fig-show-basic-plots
#| fig-cap: "A: total number of job ads posted to the Academic Jobs Wiki for Archaeology in each year, with coloured sections showing the proportion of jobs by title and rank. B: Ratio of tenure-track to non-tenure-track positions over time.The proportion of the bar above the red line reflects the number of non-tenure track positions available relative to tenure-track positions."
#JB: I have added the last sentence above because I have not seen this kind of figure in an archaeology paper, so I think we should explain it. Will reduce friction for our readers.Please confirm that I am actually correctly explaining how the figure works, however.
#JB: I would be eternally grateful if you would convert the hyphens between the timespans on the x-axis into en dashes 😬

# show figure from PNG file
knitr::include_graphics(here("analysis",
            "figures", 
            "fig-panel-per-year.png"))

@@

```{=html}
<!--Comments on Abstract and Keywords: 
1. Is “jobs listing website” the appropriate category here? I often hear these referred to as “jobs boards”—that might be a better fit?
2. Structurally, the second research domain is phrased as a question. For consistency, it would be good to provide a similar structure for the first domain, e.g. “Our analysis focuses on the XXX—have the topics, methods, or geographic regions specified in archaeological job ads changed over time?”
3. Suggested keywords: archaeology, hiring, academia, job ads-->
```

@@

The academic job market is a source of uncertainty, for both job seekers and hiring committees. Job seekers must search for ads that are well-matched to the skill sets and professional experiences that they have taken years to cultivate. Hiring committeesmust sift through tens to hundreds of applicants in the hope of finding a candidate that meets all of their needs and helps to realize<!--We should go with American spelling, no? With your Australian inclinations and my growing incorporation of Irish orthography into my writing, this paper is gonna be a mess...--> their visions of their program and the discipline. Finding the right candidate for the position—or the right position for the candidate—is akin to finding a needle in a haystack.

There is also a perception among many scholars that the last three decades have seen the academic job market become increasingly competitive. Across the social sciences and humanities, there are fewer jobs available relative to the number of PhD graduates and higher numbers of short-term positions relative to permanent positions (**Bessner and Brenes 2021; Brami et al., 2023; Kawa et al. 2019; Kelsky 2015; Mackie and Rockwell 2023**). This market saturation has arisen in tandem with increasingly complex application requirements. Many job ads now solicit specific documents in *addition* to the cover letter and CV, such as teaching, research, and diversity statements. These additional documents must be tailored for each application, making the process of applying for jobs a full-time job of its own.

In anthropology in the US the number of doctoral anthropology graduates has increased by about 70% over the past 30 years, but the number of new faculty positions has not increased proportionally [@speakman2018choosing]. New faculty positions have dwindled, in part due to the removal of the of the Age Discrimination in Employment Act (ADEA) exemption in 1994 which prohibited mandatory retirement ages in higher education (**Earle and Delpo Kulow 2015**). When combined with the institution of tenure in American academia, the ADEA exemption allowed [faculty to stay in their posts for as long as they liked] the median age for faculty in the U.S. now ranks among the highest for all professions (**Citation**). In tandem with the gradual de-investment in American higher education since the **XXX (Mirowski**) and the aftershocks of the 2008 recession, the ADEA exemption has cultivated an environment where new lines are few and far between. Among biological anthropologists, @Passalacqua_2018 found a ratio of 0.81 PhDs to job academic advertisements per year, concluding that academic positions in biological anthropology are barely at sustainable levels. This echoes findings from other fields. In biomedical sciences, for example, there is one tenure-track position in the US for approximately every 6.3 PhD graduates [@Ghaffarzadegan_2015]. **Citation of Larson et al. 2014**.

PhD graduates who do not go directly into a tenure-track faculty position after graduation often go into a series of short-term appointments as part-time or limited contract instructors, also known as adjunct or contingent faculty. Rising numbers of contingent faculty has long been a concern in US higher education [@Trevithick2010]. Data from the American Association of University Professors' 2021–22 faculty survey indicate that more than 60% of faculty positions in US universities were held by non-tenure track full time or part time contingent faculty members [@colby2022annual]. Contingent faculty positions are precarious because they provide low or no health and retirement benefits, little or nothing in the way of professional development, and lower salaries relative to tenure-track positions. PhD graduates who have a sequence of short-term academic positions are often disadvantaged financially, due to low compensation combined with expensive frequent relocation, socially due to isolation from family and community, and professionally due to being ineligible for many decision-making roles at the universities they work at [@platzerAcademicPrecarityAmerican2018].

Success in tenure-track job applications in archaeology is strongly determined by where applicants get their PhDs. Data from the 2014–2015 AnthroGuide publication of the American Anthropological Association shows that just ten out of over 100 US graduate programs produced over 30% of the graduates hired into tenure-track faculty positions [@speakman2018choosing]. Similarly, a network analysis of 1,918 faculty holding tenured or tenure-track positions at PhD-granting anthropology programs in the US in 2015 (including 506 archaeologists) found that the just fifteen graduate programs produced 52.8 percent of tenured and tenure-track positions [@kawaSocialNetworkUS2019]. This network analysis showedthat programs with large endowments and with faculty who hold prestigious awards and are widely cited by other scholars produce the majority of tenured and tenure-track faculty in anthropology. Hiring bias predicated on the prestige of specific institutions and programs is not unique to anthropology. Targeted studies of the disciplines of sociology, communication, and operations research, and industrial systems engineering show similar trends  (**Barnett et al., 2010; Castillo et al., 2018; Feeley and Tutzauer 2022; Nevin 2019**). **Broader analyses comparing X, Y and Z, and analyses while a recent analysis of XXX showed Z (Clauset et al., 2015;** [@wapmanQuantifyingHierarchyDynamics2022a]). Within anthropology, however, such studies draw attention to the paradox of a discipline committed to fighting social inequalities continuing to reproduce its own systematic inequalities through hiring practices that favor an elite minority of graduates with prestigious institutional affiliations.

One way that some hiring committees are tackling these biases towards prestigious institutions is by providing detailed instructions to applicants on how to prepare their application materials. In theory, detailed applications will allow the hiring committee to focus on evaluating the accomplishments of candidates across common categories, rather than ranking based on prestige signals in a CV, such as a name of the applicant's graduate program. This push for equity has resulted in job ads that are often highly prescriptive in the types of documents that applicants should submit. For example, in addition to a cover letter and CV, job ads in many fields now require applicants to submit short statements detailing their previous and future contributions to teaching, research, and diversity within their discipline. In a comparison of job ads from 1999–2000 and 2019–2020 published in *Anthropology News* @HelloTristesTropes2021 noticed an increase in the number of materials requested from applicants. For example, twice as many 2019–2020 job ads requested writing samples compared to 1999–2000, and nearly four times as many requested statements of teaching philosophy. In their review of the 'worst job ads of 2021' @dennisWorstAnthroJob2022 reported two job ads with particularly extreme requirements<!--Are you opposed to naming and shaming these institutions, which were Oberlin College and Grinnell College? I applied for one of these positions, and have no issue referring to them by name...-->, which respectively requested nine and thirteen documents from applicants. This high burden on applicants disproportionately favors people with more time and financial resources to prepare the required materials.

A notable requirement that has emerged in recent years is a diversity statement, where the applicant describes their knowledge of, prior contributions to, and future goals for advancing diversity, equity, inclusion. This requirement has been much-debated largely due to an experiment in 2016–2022 at several University of California campuses that used diversity statements as the first cut for selecting candidates for tenure-track faculty positions [@soucek2021diversity]. Only candidates that scored highly on their diversity statements would have the rest of their application evaluated. This experiment generated intense and widespread public debate about the merits and risks of requiring and using diversity statements. These debates drew attention to the challenges of hiring faculty from underrepresented minorities, and resulted in a variety of approaches to evaluating diversity, even leading some universities to entirely omit requirements for diversity statements in job applications [@guidenAreDiversityStatements].

**While attention to equity in job market materials is commendable, work on hiring practices suggests ZZZ. For example, Warren 2019 has shown XXX. Rivera, in a study of hiring practices at YYY, has likewise demonstrated ZZZ. Somewhere in this section I want to discuss the results from sociology, specifically the Warren 2019 piece and the Rivera 2017 piece on hiring practices.**

**Cite Rogge 1976.**

@@

Our primary data source is the Archaeology Academic Jobs Wiki<!--Check if Kelsky has anything too say about this. I also think there have been some articles about the pros and cons of the poli sci or Econ versions of this.-->. Originating in 2007, this is a set of freely accessible web pages that anyone can edit (anonymously or with a free user account) hosted by Fandom, a for-profit company. The Archaeology pages are part of the Academic Jobs Wiki, which coordinates similar collaboratively-edited resources for around 40 academic disciplines. The coordinators and contributors are nearly all anonymous or pseudonymous. Typically, contributors copy and paste the text of job ads from other sources, such as the *Chronicle of Higher Education*, *Higher Ed Jobs*, and university websites, into the wiki, collecting <!--consolidating? centralizing?--> ads originally posted in numerous different locations. Other contributors then edit the web page to add comments below an ad to share relevant information based on their experience in applying for that position, such as a tally of how many people have applied, the dates of events such as requests for more materials, interviews, offer made, rejection notices, etc. Contributors also edit the page to ask and answer questions about the positions and the application process. These comments make the Academic Jobs Wiki a unique resource for timely and specific information for job-seekers about positions they are interested in, and one of the most important internet resources for the academic job market. Because of its reputation for aggregating ads from diverse sources and rapidly-updated information that is not available elsewhere, the Academic Jobs Wiki has a large community of users that keep it updated and accurate has become an authoritative data source for studies of hiring trends in academia [e.g. @Passalacqua_2018; @musial2018five] and a widely recommended resource [e.g. @lightfoot2021preparing].

For each tenure-track job advertised on the Archaeology Academic Jobs Wiki during 2013-2023, we read the text and recorded the name of the hiring institution, the title of the position, and exact words and phrases from the ad about the topical, geographic, and methods foci on the position into a Google form. The topical focus is what we understood as the primary intellectual focus of the position—**examples of topical foci included X,Y, and Z**. The geographic focus is the region of the world that the ideal candidate has scholarly expertise on, **for example, X, Y, or Z**. The methods focus is the data-generating sub-field of archaeology that is mentioned in the ad. **Examples of methods used in this study include X, Y, or Z.** <!--It might be helpful to make this into a summary table with examples of all foci generated by the study, and then one showing the simplified 10–15 topics. I understand this is in the supplementary tables already.--> We also recorded the type and number of documents requested in each ad (e.g. cover letter, CV, statements on research, teaching, diversity<!-- Were these all relatively easy to recognise? I’ve applied to places like Penn State, for example, that do a stupid combined research/teaching/diversity statement, just to make applicants do extra work, seemingly.-->, syllabi, course descriptions, writing samples, transcripts) and how many names/letters of recommenders were requested in the ad.

After completing primary data collection, we studied the topical, geographic, and methods foci of each ad and collaboratively and manually reduced the variation in the raw data into 10-15 categories appearing in 20 (for topics and geography) or 10 (for methods) or more job ads to simplify analysis and visualisation. This means that some topics, such as gender (mentioned in `r n_jobs_mentioning_gender` ads) do not appear in our results because of their rarity in the job ads. Full details of the category reduction, showing the mapping between phrases in the job ads and our categories, are in our Supplementary Materials. <!--
I think it would be much easier for our readers to parse this info as a table, unless you have any objection. This would also help us rein in our word count after all my contributions-->Our topic categories were: American archaeology, Ancient Europe and Mediterranean, Archaeological science, Archaeological theory, Biological anthropology, Complex societies, Digital archaeology, Environmental archaeology, Evolutionary anthropology, Indigenous and Historical archaeology, North Mesoamerican Archaeology, Pleistocene archaeology, and Public archaeology Our geographic categories were: Africa, Americas, Asia & India, Canada & Arctic, Europe, Mediterranean, Meso- & South America, Near East, Oceania, Midwest US, Northeastern US, Southeast US, Southwest US, and Western US. Our methods categories were: Archaeobotany, Archaeometry, Bioarchaeology, Ceramic analysis, Computational and Digital archaeology, Geoarchaeology, Landscape analysis, Lithic analysis, Material culture analysis, and Zooarchaeology. Ads could have multiple or none of these three foci, and some of the foci overlap. Some topics include geographic regions because this is how they are typically understood by archaeologists. For example, Mesoamerican archaeology is understood to refer to a specific time period and geographic region. Similarly, we recorded digital archaeology we recorded as both a method (when a job ad had a clearly distinct topical focus, such as historic archaeology) and a topic (when there were no other topics mentioned in the job ad). While these categorical overlaps can make the data challenging to interpret, in our view this approach reflects the complex realities of how search committees express their needs in searching for new faculty. Acknowledgement of overlaps also produces new insights into hiring dynamics through revealing intersections between different foci.

@@

We collected data from `r total_number_of_ads_in_our_sample` ads for tenure-track jobs in archaeology posted during 2013-2023. We focus our analysis here on the `r total_number_of_us_ads_in_our_sample` ads for positions at US universities because <!--Ben,we should provide the rationale for this, I don't remember what it was-->. @fig-show-basic-plots shows the count of ads for each year. Assistant Professor jobs are consistently the most common title and rank of positions advertised, while open rank or full professor are the least frequent. The ratio of tenure-track to non-tenure track positions is generally well above one; in other words, this data set is dominated by tenure-track positions . Only academic year 2013-2014 had more non-tenure track positions than tenure-track positions <!--I know that this is a pedantic complaint, but we should be consistent about whether or not we hyphenate tenure-track-->, which was followed by an upward trend peaking at 2018-2019 and then declining again into the present.

@@

```{r}
#| label: fig-show-map-of-hiring-institution-and-cc
#| fig-cap: "A: Frequency of hiring institution by Carnegie classification. B: Inset shows map of the United States showing the count of tenure-track job ads posted by all insititutions in each state during 2013 -- 2023"
#JB: So many positions in New York! Why don’t I have a job in New York?!
# JB: This figure is great. We should make sure it works for colour-blind readers.

@@

Panel B of @fig-show-map-of-hiring-institution-and-cc shows the geographic distribution of the hiring institutions. California posted almost twice as many job ads as the next most active states. After California, the states that posted the most ads during 2013-2023 include New York, Texas, and Pennsylvania, and Florida. These top five states also correspond to the top five most populous US states, suggesting that rates of hiring are approximately proportional to population density. These top five states for job ads are also the top five states with highest number of degrees awarded in Anthropology (U.S. Department of Education nd). Similarly, the lowest counts of job ads were observed in states with the lowest populations: North Dakota, South Dakota, Alaska, and Nebraska. No institutions in Montana posted a job ad during this period. The implication here is that job-seekers who are able to relocate to populous areas will have more employment options.<!--Yes, nicely put-->

@@

#JB: Interesting that Near East is on the up and up. You know what I am going to say about the timespans in the x-axis…

@@

We recorded all geographic regions mentioned in the text where the successful applicant should have expertise and be research active. Our analysis focuses on those locations mentioned in 20 or more ads. Overall, American locations dominate. Panel A of @fig-show-geo-trends shows that a single region of the US, the Southwest, occurs in more job ads than every other part of the world except for the Mediterranean. <!-- Someone in another field recently asked me what archaeologists wear to conferences, and I said “bolo ties”. This is why.-->The Southwest includes Arizona and New Mexico, with portions of California, Colorado, Nevada, Oklahoma, Texas, and Utah. It is archaeologically significant as the home of the Ancestral Pueblo, Hohokam, and Mogollon peoples who practiced irrigation agriculture and lived in relatively large settlements, compared to other regions of the US. The area was later occupied by the Navajo, Ute, Southern Paiute, Hopi and Zuni, also in relatively large numbers (**Summary citation(s), preferably including an Indigenous archaeologist from the region)**. The Mediterranean is prominent because it is the location that is often mentioned in job ads focused on classical archaeology (i.e. archaeology of Bronze and Iron Age Italy and Greece).

Demand for jobs focusing on the Americas has been generally high over the past decade, with a peak in 2019-2020, and a subsequent decrease. Demand for jobs focusing on Africa was very low until 2019-2020, peaking in 2020-2021. The proportion of ads with a geographic focus on the Mediterranean has varied substantially, peaking in 2016 and experiencing a nadir in 2019, showing an inverse pattern to the Americas. Asia and India, the Near East, and Europe rarely occur as a geographical focus in job ads at US institutions. Asia and India, Africa, and the Americas appear correlated with each other, while the Near East and Mediterranean are correlated in an opposite trend.

@@

Landscape archaeology, encompassing GIS and remote sensing, has remained prominent compared to other methods [@fig-show-metho-trends]. <!--Might be worth exploring whether this is a skillset more frequently sought after in CRM and/or other industries—is this a transferable skill thing?--> Methods focused on a specific element of the archaeological record, such as Lithic analysis, Zooarchaeology and Ceramics are among the least frequently mentioned in job ads. Instead, the more popular methods are those that are relevant to multiple elements of the archaeological record (e.g. Archaeobotany encompasses macroscopic and microscopic plant remains; Bioarchaeology may include skeletal analysis, isotopes, proteins, etc.).

Landscape archaeology, although dominant, has fluctuated over the years and has been on a downtrend since 2018-2019. Computational and digital archaeology is the second most represented method, showing an overall increasing trend, particularly since 2020-2021. <!--Congratulations--> Archaeobotany shows a strong cyclical trend, rising, falling, and then rising again over our study period. Archaeometry and Geoarchaeology have maintained a relatively low but steady presence in job ads, peaking in 2017-2018 and 2018-2019 and declining thereafter. Lithic analysis and Zooarchaeology are also mentioned relatively infrequently in job ads and show an inverse correlation with each other after 2018-2019.

@@

#JB: Please. The x-axis. My fragile mental state.

#JB: Given that not all job ads are categorised by all foci, can we provide an overall n for Figure 3 for and 5, either in the captions or in the figures themselves? I assume that 20% of our “methods” sample is not the same as 20% of our “topics” sample.

@@

The most frequently mentioned topic in the job ads is Environmental archaeology [@fig-show-topi-trends]. This category encompasses such phrases as human-environmental dynamics, interaction between humans and their environments, environmental change, climate change, historical ecology, ecological knowledge,human ecology, and ecological systems. Public archaeology is the second most frequent topic overall, which included phrases such as cultural resource management, cultural heritage, heritage studies, museum studies, human rights, community engaged, historic preservation, social justice, community-based, repatriation, and community-engaged archaeology. The least common topics in our sample are Pleistocene archaeology, e.g. human origins, hunter-gatherer archaeology, and digital archaeology.

In the years 2019-2020 and 2020-2021 there are striking changes in the popularity of topics in job ads. Indigenous and historical archaeology (which includes archaeology of enslaved people <!--African diaspora archaeology? We should be careful to use the appropriate language here-->) became the most popular topic at this time, rising from being one of the least popular topics from 2012-2017. Conversely, archaeological science, which was popular during 2012-2017, was rarely mentioned in job ads during 2019-2021. Ancient Europe and the Mediterranean, which was an infrequently mentioned topic for the entire study period, virtually disappeared from job ads during 2019-2021. Mentions of complex societies in ads decrease at a steady rate, while mentions of public archaeology consistently increase over the study period. The topic of Environmental Archaeology remains consistently <!--We should be careful of the word consistent as we are relying on it a bit heavily throughout the text. Here, could we replace with "popular" or "stable"?-->high over time.
